Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
work on intro
@mawilliams99 , FYI, I'm working on this today.

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
author:
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "1,2" 
  - given_name: Ben 
    surname: Bond-Lamberty
    affiliation: "3" 
  - given_name: Susan 
    surname: Cook-Patton
    affiliation: "4" 
  - given_name: Helene
    surname: Muller-Landau
    affiliation: "5" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "6" 
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,5 *"
    email: teixeirak@si.edu
    corresponding: true
affiliation:
  - code: 1
    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
  - code: 2
    address: School of Geography, University of Leeds, Leeds, UK
  - code: 3
    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
  - code: 4
    address: The Nature Conservancy; Arlington VA 22203, USA
  - code: 5
    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
  - code: 6
    address: CIRAD, Montpellier, France
abstract: |
  Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
  Yet, the efficacy of greenhouse gas (GHG) mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
  Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
  Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector GHG emissions and removals.
  We begin by reconciling terminology and mapping ForC fields into EFDB.
  This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
  At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
  (Some analysis of representation/ gaps.)
  <!-- -->
  In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.

@@

Forests are critical to our management of greenhouse gases. 
The influence the forests have on atmospheric carbon has made forest data increasingly valuable.

**Forests play a central role in international plans for climate change mitigation [@unfccc_adoption_2015].**
As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
Effective forest-based climate change mitigation requires solid scientific data [@anderson-teixeira_effective_2022].

**The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas (GHG) inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, (REFS for older guidelines), @ipcc_2019_2019].**
Under this guidance, GHG inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for forest land ... [@ipcc_agriculture_2006; @ipcc_2019_2019a].
This guidance has improved as more of the relevant underlying data has become available [@requenasuarez_estimating_2019], but there remains room for continuous improvement as the science advances.
For example, @cook-patton_mapping_2020 *(summarize IPCC comparison results)*. 
Another example is @cuni-sanchez_high_2021.
*Moreover, it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.*
To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.

**To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished GHG reductions, or as a basis of comparison for external parties to evaluate these inventories. 
<!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->

@@

The UNFCCC requires GHG reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*]. 
This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).

>>>>

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
author:
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "1,2" 
  - given_name: Ben 
    surname: Bond-Lamberty
    affiliation: "3" 
  - given_name: Susan 
    surname: Cook-Patton
    affiliation: "4" 
  - given_name: Helene
    surname: Muller-Landau
    affiliation: "5" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "6" 
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,5 *"
    email: teixeirak@si.edu
    corresponding: true
affiliation:
  - code: 1
    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
  - code: 2
    address: School of Geography, University of Leeds, Leeds, UK
  - code: 3
    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
  - code: 4
    address: The Nature Conservancy; Arlington VA 22203, USA
  - code: 5
    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
  - code: 6
    address: CIRAD, Montpellier, France
abstract: |
  Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
  Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
  Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
  Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
  We begin by reconciling terminology and mapping ForC fields into EFDB.
  This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
  At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
  (Some analysis of representation/ gaps.)
  <!-- -->
  In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.

@@

Forests are critical to management of atmospheric concentrations of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> [forests only, @harris_global_2021].
This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al. Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@refs], and on human conservation, restoration, and management of forests [@refs].

Reflecting their strong influence on Earth's climate, forests play a central role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [@refs].
One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning, underlain by solid scientific data [@anderson-teixeira_effective_2022].

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones. 
In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.
To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.

**To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
<!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->

@@

The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*]. 
This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
work on intro
@mawilliams99 , FYI, I'm working on this today.

\begin{abstract}
Forests are critical for climate change mitigation and consitute a
substantial portion of planned emissions reductions under the 2015 Paris
Agreement. Yet, the efficacy of greenhouse gas (GHG) mitigation planning
and reporting is dependent upon the quality of available emission
factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published,
yet are not readily accesible to the practitioners compiling national
greenhouse gas inventories. Many of these data have, however, been
compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector GHG emissions and
removals. We begin by reconciling terminology and mapping ForC fields
into EFDB. This process required some updates to the ForC database
structure, leading to the release of a new version of ForC (v4.0;
described here). At the time of writing, ForC contained \#\# values that
would qualify for inclusion in the EFDB, \#\# of which have been
transferred to date. (Some analysis of representation/ gaps.) In the
future, forest C estimates in EFDB can be improved through targetted
research to fill critical gaps, reporting of information required by
IPCC, and continued submission of data from scientific publications to
the EFDB.
\end{abstract}

@@

Forests are critical to our management of greenhouse gases. The
influence the forests have on atmospheric carbon has made forest data
increasingly valuable.

\textbf{Forests play a central role in international plans for climate
change mitigation \citep{unfccc_adoption_2015}.} As of 2016,
forest-based mitigation accounted for 26\% of total planned greenhouse
gas mitigation within Nationally Determined Contributions under the
Paris Agreement \citep{grassi_key_2017}. Effective forest-based climate
change mitigation requires solid scientific data
\citep{anderson-teixeira_effective_2022}.

\textbf{The International Panel on Climate Change (IPCC) provides
guidance for national greenhouse gas (GHG) inventories for reporting to
the United Nations Framework Convention on Climate Change \citep[UNFCCC,
(REFS for older guidelines),][]{ipcc_2019_2019}.} Under this guidance,
GHG inventories include all managed land, including most of the world's
forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
include specific instructions for accounting for forest land \ldots{}
\citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This guidance has
improved as more of the relevant underlying data has become available
\citep{requenasuarez_estimating_2019}, but there remains room for
continuous improvement as the science advances. For example,
\citet{cook-patton_mapping_2020} \emph{(summarize IPCC comparison
results)}. Another example is \citet{cuni-sanchez_high_2021}.
\emph{Moreover, it is useful for those compiling national greenhouse gas
inventories to have access to locally-specific information, when
available.} To improve the data accessible for C accounting, the IPCC
created the Emission Factor Database (EFDB;
https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
recognized library of emission factors and other parameters that can be
used for estimating greenhouse gas emissions and removals.

\textbf{To ensure that planned emissions reductions are realistic,
high-quality estimates of forest C stocks and fluxes must be publicly
accessible.} The EFDB is intended as a recognized library, and can be
used both for efforts to tally a nation's intended or accomplished GHG
reductions, or as a basis of comparison for external parties to evaluate
these inventories.

@@

The UNFCCC requires GHG reporting for all managed lands in a country,
where management is defined as ``human interventions and practices have
been applied to perform production, ecological or social functions''
{[}\emph{IPCC 2006 full report REF}{]}. This definition is applied
differently across countries, and is not clearly defined by the majority
of governments \citep{ogle_delineating_2018}. Given this, and because
the IPCC definition of management does not necessarily match that which
would be reported in scientific publications and hence in ForC, we do
not transfer any classification of land management status from ForC to
the EFDB, but do provide auxiliary info that may be useful in making
this determination (e.g., geographical location).

>>>>

\begin{abstract}
Forests are critical for climate change mitigation and consitute a
substantial portion of planned emissions reductions under the 2015 Paris
Agreement. Yet, the efficacy of greenhouse gas mitigation planning and
reporting is dependent upon the quality of available emission factors
data, including forest carbon (C) stocks and changes therein. Tens of
thousands of relevant forest C estimates have been published, yet are
not readily accesible to the practitioners compiling national greenhouse
gas inventories. Many of these data have, however, been compiled in the
Global Forest C database (ForC; https://forc-db.github.io/) and stand to
be of value to greenhouse gas accounting if made available through the
Emission Factor Database (EFDB) of the International Panel on Climate
Change (IPCC). Here, we develop and document a process for
semi-automated transfer of data from ForC into the EFDB, assess the data
available and transferred to date, and provide recommendations for
improving forest data collection, analysis, and reporting to improve
accounting of forest-sector greenhoouse gas emissions and removals. We
begin by reconciling terminology and mapping ForC fields into EFDB. This
process required some updates to the ForC database structure, leading to
the release of a new version of ForC (v4.0; described here). At the time
of writing, ForC contained \#\# values that would qualify for inclusion
in the EFDB, \#\# of which have been transferred to date. (Some analysis
of representation/ gaps.) In the future, forest C estimates in EFDB can
be improved through targetted research to fill critical gaps, reporting
of information required by IPCC, and continued submission of data from
scientific publications to the EFDB.
\end{abstract}

@@

Forests are critical to management of atmospheric concentrations of the
greenhouse gas carbon dioxide (CO\textsubscript{2}), and thereby climate
change. In recent decades, CO\textsubscript{2} uptake by forests,
woodlands, and savannas has exceeded releases from deforestation and
other severe disturbances, resulting in a net carbon CO\textsubscript{2}
sink of \textasciitilde0.88 Gt C yr\textsuperscript{-1} \citep[all
biomes with trees,][]{xu_changes_2021} to \textasciitilde1.6 Gt C
yr\textsuperscript{-1} \citep[forests only,][]{harris_global_2021}. This
has offset an estimated 10\% to 18\% of anthropogenic
CO\textsubscript{2} emissions from fossil fuels and cement
\citep{xu_changes_2021, harris_global_2021}, dramatically slowing the
pace of atmospheric CO\textsubscript{2} accumulation and climate change.
Going into the future, the fate of this important CO\textsubscript{2}
sink is highly uncertain, depending both upon forest responses to
climate change, which are likely to reduce the sink strength
\citep{refs}, and on human conservation, restoration, and management of
forests \citep{refs}.

Reflecting their strong influence on Earth's climate, forests play a
central role in international plans for climate change mitigation under
the Paris Agreement \citep{unfccc_adoption_2015}. Forest conservation,
reforestation, and improved sustainable management all have significant
-- and relatively cost-effective -- potential as climate change
mitigation options, with conservation and reforestation having the
fourth and fifth largest net emission reduction potentials or all
mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
mitigation accounted for 26\% of total planned greenhouse gas mitigation
within Nationally Determined Contributions under the Paris Agreement
\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
mitigation initiatives do not always correspond to actual emission
reductions through on-the-ground implementation \citep{refs}. One
critical need for ensuring that forest-based climate change mitigation
initiatives are effective is realistic planning, underlain by solid
scientific data \citep{anderson-teixeira_effective_2022}.

The International Panel on Climate Change (IPCC) provides guidance for
national greenhouse gas inventories for reporting to the United Nations
Framework Convention on Climate Change
\citep[UNFCCC,][]{ipcc_2019_2019}. Under this guidance, greenhouse gas
inventories include all managed land, including most of the world's
forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
include specific instructions for accounting for greenhouse gas (mainly
CO\textsubscript{2}) exchanges between forest land and the atmosphere
\citep{ipcc_agriculture_2006, ipcc_2019_2019}. This guidance has
improved over the years as more of the relevant underlying data has
become available
\citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC
guidelines, \citet{cook-patton_mapping_2020} found that the latest
default rates may underestimate rates of C accumulation in regrowth
forests by 32\% on average and fail to capture eight-fold variation
within ecozones. In addition, \citet{cuni-sanchez_high_2021} found that
aboveground C stocks in mature African tropical montane forests were
two-thirds higher than the IPCC default values for these forests. This
rapid evolution of scientific information on the climate mitigation
potential of forests is beneficial to climate mitigation efforts, but
requires improved mechanisms for communicating the latest information
from scientific researchers to the practitioners who need reliable
estimates for greenhouse gas mitigation planning. Moreover, high
variability of forest C cycling within ecozones
\citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
useful for those compiling national greenhouse gas inventories to have
access to locally-specific information, when available. To improve the
data accessible for C accounting, the IPCC created the Emission Factor
Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which
is intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals.

\textbf{To ensure that planned emissions reductions are realistic,
high-quality estimates of forest C stocks and fluxes must be publicly
accessible.} The EFDB is intended as a recognized library, and can be
used both for efforts to tally a nation's intended or accomplished
greenhouse gas reductions, or as a basis of comparison for external
parties to evaluate these inventories.

@@

The UNFCCC requires greenhouse gas reporting for all managed lands in a
country, where management is defined as ``human interventions and
practices have been applied to perform production, ecological or social
functions'' {[}\emph{IPCC 2006 full report REF}{]}. This definition is
applied differently across countries, and is not clearly defined by the
majority of governments \citep{ogle_delineating_2018}. Given this, and
because the IPCC definition of management does not necessarily match
that which would be reported in scientific publications and hence in
ForC, we do not transfer any classification of land management status
from ForC to the EFDB, but do provide auxiliary info that may be useful
in making this determination (e.g., geographical location).
